Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-07-24 paper/paper.Rmd
Fix footnote and alg name

The inner product between two vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $|u|=\sqrt{\langle u,u \rangle_g}$ and the angle between two vectors $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$ in the tangent space.
[^1]: Here the Einstein notation is used where superscripts denote summation over $i$ and $j$.

The line element and volume element of the full manifold or a subset of the manifold can also be computed from $g$. The arc length of a curve $c$ along the manifold is defined as
$$
l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
$$ 
where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and $c(t)$ is a function that maps $[a,b]$ to $M$. While the volume of $V\subset M$ is computed by 
$$
Vol(V)=\int_V \sqrt{|\det g|} dx^1\dots dx^d.
$$ 
Both the concepts of distance and volume are relevant to kernel density estimation as we have shown in Equation \@ref(eq:volumedensityratio), \@ref(eq:volumedensityest), \@ref(eq:dgestimate), and \@ref(eq:vbs).

@@

In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm. For visualization purposes, \autoref{twodgaussian} presents a 2-D meta data example. We first simulate the data of size $N=2,000$ from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of $500$ points. Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP. The embedded dimension is set as $d=2$, the same as the meta data dimension.
This enables us to compare the manifold learning embedding with the true meta data. We could now apply Algorithm \@ref(alg:vkderm) to get the density estimates of all data points and further detect anomalies. As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size $N=2,000$ embedded in a 100-D space and the corresponding embedding dimension is $d=5$.

@@

In this section, we first calculated the same empirical distributions of the $336$ half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in @Cheng2021-ex in the statistical manifold learning to get the 2-D embedding of all households. Algorithm \@ref(alg:vkderm) is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers. The data processing steps have been clearly
stated in the application section of @Cheng2021-ex and they are skipped here. Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of the week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}. However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.

>>>>

The inner product between two vectors $u,v \in T_{\pmb{p}}M$, given by $\langle u,v \rangle_g=g_{ij}u^iv^j$[^1], can be used to define some geometric quantities, such as the vector norm $|u|=\sqrt{\langle u,u \rangle_g}$ and the angle between two vectors $\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}$ in the tangent space.
The line element and volume element of the full manifold or a subset of the manifold can also be computed from $g$. The arc length of a curve $c$ along the manifold is defined as
$$
l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
$$ 
where $(x^1,\dots,x^d)$ are the coordinates of chart $(U,x)$ and $c(t)$ is a function that maps $[a,b]$ to $M$. While the volume of $V\subset M$ is computed by 
$$
Vol(V)=\int_V \sqrt{|\det g|} dx^1\dots dx^d.
$$ 
Both the concepts of distance and volume are relevant to kernel density estimation as we have shown in Equation \@ref(eq:volumedensityratio), \@ref(eq:volumedensityest), \@ref(eq:dgestimate), and \@ref(eq:vbs).

[^1]: Here the Einstein notation is used where superscripts denote summation over $i$ and $j$.

@@

In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm. For visualization purposes, \autoref{twodgaussian} presents a 2-D meta data example. We first simulate the data of size $N=2,000$ from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of $500$ points. Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP. The embedded dimension is set as $d=2$, the same as the meta data dimension.
This enables us to compare the manifold learning embedding with the true meta data. We could now apply Algorithm \@ref(alg:kdeanomaly) to get the density estimates of all data points and further detect anomalies. As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size $N=2,000$ embedded in a 100-D space and the corresponding embedding dimension is $d=5$.

@@

In this section, we first calculated the same empirical distributions of the $336$ half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in @Cheng2021-ex in the statistical manifold learning to get the 2-D embedding of all households. Algorithm \@ref(alg:kdeanomaly) is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers. The data processing steps have been clearly
stated in the application section of @Cheng2021-ex and they are skipped here. Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of the week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}. However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.

### 2022-07-24 paper/paper.tex
Fix footnote and alg name

The inner product between two vectors \(u,v \in T_{\pmb{p}}M\), given by \(\langle u,v \rangle_g=g_{ij}u^iv^j\){[}\^{}1{]}, can be used to define some geometric quantities, such as the vector norm \(|u|=\sqrt{\langle u,u \rangle_g}\) and the angle between two vectors \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\) in the tangent space.
{[}\^{}1{]}: Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\).

The line element and volume element of the full manifold or a subset of the manifold can also be computed from \(g\). The arc length of a curve \(c\) along the manifold is defined as
\[
l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
\]
where \((x^1,\dots,x^d)\) are the coordinates of chart \((U,x)\) and \(c(t)\) is a function that maps \([a,b]\) to \(M\). While the volume of \(V\subset M\) is computed by
\[
Vol(V)=\int_V \sqrt{|\det g|} dx^1\dots dx^d.
\]
Both the concepts of distance and volume are relevant to kernel density estimation as we have shown in Equation \eqref{eq:volumedensityratio}, \eqref{eq:volumedensityest}, \eqref{eq:dgestimate}, and \eqref{eq:vbs}.

@@

In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm. For visualization purposes, \autoref{twodgaussian} presents a 2-D meta data example. We first simulate the data of size \(N=2,000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points. Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP. The embedded dimension is set as \(d=2\), the same as the meta data dimension.
This enables us to compare the manifold learning embedding with the true meta data. We could now apply Algorithm \ref{alg:vkderm} to get the density estimates of all data points and further detect anomalies. As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size \(N=2,000\) embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).

@@

In this section, we first calculated the same empirical distributions of the \(336\) half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in \textcite{Cheng2021-ex} in the statistical manifold learning to get the 2-D embedding of all households. Algorithm \ref{alg:vkderm} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers. The data processing steps have been clearly
stated in the application section of \textcite{Cheng2021-ex} and they are skipped here. Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of the week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}. However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.

>>>>

The inner product between two vectors \(u,v \in T_{\pmb{p}}M\), given by \(\langle u,v \rangle_g=g_{ij}u^iv^j\)\footnote{Here the Einstein notation is used where superscripts denote summation over \(i\) and \(j\).}, can be used to define some geometric quantities, such as the vector norm \(|u|=\sqrt{\langle u,u \rangle_g}\) and the angle between two vectors \(\cos{\theta}=\frac{\langle u,v \rangle_g}{|u| |v|}\) in the tangent space.
The line element and volume element of the full manifold or a subset of the manifold can also be computed from \(g\). The arc length of a curve \(c\) along the manifold is defined as
\[
l(c)=\int_a^b \sqrt{g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt}} dt,
\]
where \((x^1,\dots,x^d)\) are the coordinates of chart \((U,x)\) and \(c(t)\) is a function that maps \([a,b]\) to \(M\). While the volume of \(V\subset M\) is computed by
\[
Vol(V)=\int_V \sqrt{|\det g|} dx^1\dots dx^d.
\]
Both the concepts of distance and volume are relevant to kernel density estimation as we have shown in Equation \eqref{eq:volumedensityratio}, \eqref{eq:volumedensityest}, \eqref{eq:dgestimate}, and \eqref{eq:vbs}.

@@

In this section, we examine two scenarios for both low and high dimensions to test our proposed algorithm. For visualization purposes, \autoref{twodgaussian} presents a 2-D meta data example. We first simulate the data of size \(N=2,000\) from a mixture of four Gaussian kernels with the same covariance but different means, each consisting of \(500\) points. Different mapping functions are then applied to the 2-D meta data to be mapped in a 3-D feature space, which gives the higher-dimensional input for different manifold learning algorithms, including ISOMAP, LLE, Laplacian Eigenmaps, t-SNE, and UMAP. The embedded dimension is set as \(d=2\), the same as the meta data dimension.
This enables us to compare the manifold learning embedding with the true meta data. We could now apply Algorithm \ref{alg:kdeanomaly} to get the density estimates of all data points and further detect anomalies. As a high-dimensional example, the second simulation in \autoref{fivedgaussian} is based on a 5-D meta data of size \(N=2,000\) embedded in a 100-D space and the corresponding embedding dimension is \(d=5\).

@@

In this section, we first calculated the same empirical distributions of the \(336\) half-hourly periods of the week for each household, and apply the total variation distance estimator proposed in \textcite{Cheng2021-ex} in the statistical manifold learning to get the 2-D embedding of all households. Algorithm \ref{alg:kdeanomaly} is then used to obtain density estimates with the pointwise variable Riemannian metric as the bandwidth matrix and detect outliers. The data processing steps have been clearly
stated in the application section of \textcite{Cheng2021-ex} and they are skipped here. Unlike the simulations in \autoref{simulation}, we know nothing about the true density of the electricity distributions for all periods of the week and all households, so it is impossible to compare the estimated densities with the true meta data density as in \autoref{fig:fivedisomapden}. However, we could generate all the density estimates with the existing KDE method with fixed bandwidth, which is an optimal method for density estimation, and compare the densities from our proposed method with them.
